HyperLogLog: Exponentially Bad in Adversarial Settings

### 2021-05-05 docs/HLL_exp_bad/main.tex
updates after feedback from Mia and Anu

So far, we have treated attacks in an abstract way, working with an off-the-shelf implementation of HLL. We now illustrate what effect such attacks might have on applications relying on HLL for cardinality estimation. Specifically, we analyse in detail a recent paper~\cite{flexswitch} from the networking systems community.

The authors of~\cite{flexswitch} propose to use HLL for recording traffic flow statistics in datacenter network switches. The idea is to use HLL so as to reduce the amount of memory needed in the switches. The resulting switches are called \emph{flexswitches} in~\cite{flexswitch}. In the main example considered in~\cite[Section 3.1]{flexswitch}, each flow is identified by a 4-tuple consisting of source IP address, destination IP address, source port, and destination port. Various parameters for the HLL sketch to be used are then explored in~\cite{flexswitch}, trading memory consumed for accuracy of estimation of the number of concurrent flows. The paper also proposes to use the low range correction method of Fig.~\ref{fig:hll}. At no point does the discussion in~\cite{flexswitch} consider adversarial input (it does mention worst-case performance but only in passing, and no indication is given of how to handle this). It is clear that, by modifying the 16-bit source port (for example), an adversary could fool the switch into undercounting the flows the adversary is responsible for producing. Depending on how the flow count is used, this could lead to undetected DoS conditions, mis-provisioning of network capacity, or under-billing of the adversary for its traffic. In Appendix~\ref{sec:flexswitch} we provide further analysis of~\cite{flexswitch}, showing that each proposed use of HLL can be subverted by an S2 adversary.

>>>>

So far, we have treated attacks in an abstract way, working with an off-the-shelf implementation of HLL. We now illustrate what effect such attacks might have on applications relying on HLL for cardinality estimation. Specifically, we analyse a recent paper~\cite{flexswitch} from the networking systems community.

The authors of~\cite{flexswitch} propose to use HLL for recording traffic flow statistics in datacenter network switches. The idea is to use HLL so as to reduce the amount of memory needed in the switches. The resulting switches are called \emph{flexswitches} in~\cite{flexswitch}. In the main example considered in~\cite[Section 3.1]{flexswitch}, each flow is identified by a 4-tuple consisting of source IP address, destination IP address, source port, and destination port. Various parameters for the HLL sketch to be used are then explored in~\cite{flexswitch}, trading memory consumed for accuracy of estimation of the number of concurrent flows. The paper also proposes to use the low range correction method of Fig.~\ref{fig:hll}. At no point does the discussion in~\cite{flexswitch} consider adversarial input (it does mention worst-case performance but only in passing, and no indication is given of how to handle this). It is clear that, by modifying the 16-bit source port (for example), an adversary could fool the switch into undercounting the flows the adversary is responsible for producing. Depending on how the flow count is used, this could lead to undetected DoS conditions, mis-provisioning of network capacity, or under-billing of the adversary for its traffic. In Appendix~\ref{sec:flexswitch} we provide further analysis showing that each use of HLL proposed in~\cite{flexswitch} can be subverted by an S2 adversary.
